# Summary of Injury and Damage from Severe Weather #

## Synopsis ##
Severe weather causes injury, death, and economic damage across the U.S. every year.  In the following analysis, we categorize the events listed in the NOAA Severe Weather data, and we summarize the damages caused by those broad categories.  Finding that tornadoes account for a comparatively large amount of damages, we investigate tornado damages further.  We find that over time, injuries and fatalities caused by tornadoes in the U.S. have decreased, but that these findings can be skewed easily by extreme events.  We also found that over time, the property and crop damages caused by tornadoes have increased.  This result is questionable, as the damages were not adjusted for inflation.
## Data Processing ##

We import the data, assuming the **.csv** file is located in the working directory.
```{r}
setwd("C:\\Users\\Chris\\Documents\\GitHub\\ReproducibleResearchProject2\\ReproducibleResearchProject2")
df <- read.csv("repdata-data-StormData.csv")
```
Next we extract only those variables of interest - injury, fatality, property damage, and crop damage - together with the factor recording event types, and variables recording the date of the event and the state the event report originated from.  
```{r}
x <- df[,c("FATALITIES","INJURIES","EVTYPE","PROPDMG","PROPDMGEXP","CROPDMG","CROPDMGEXP","BGN_DATE","STATE")]
x$EVTYPE <- as.factor(x$EVTYPE)
x$DATE <- lapply(x$BGN_DATE,function (z) as.Date(strsplit(as.character(z)," ")[[1]][1],"%m/%d/%Y"))
x$YEAR <- as.factor(sapply(x$DATE,function(z) format(z,"%Y")))
x$MONTH <- as.factor(sapply(x$DATE,function(z) format(z,"%b")))
x$DATE <- NULL
```
Because many recorded event types resulted in no injury or damage, we now filter our records to include only those event types which resulted in one of the four categories of injury or damage.
```{r}
y1<-by(x$FATALITIES,x$EVTYPE,sum)
y2<-by(x$INJURIES,x$EVTYPE,sum)
y3<-by(x$PROPDMG,x$EVTYPE,sum)
y4<-by(x$CROPDMG,x$EVTYPE,sum)
examine <- levels(x$EVTYPE)[which(y1>0 | y2>0 | y3>0 | y4>0)]
x <- x[x$EVTYPE %in% examine,]
```
Also, the exponents for property damage (**PROPDMGEXP**) and crop damage (**CROPDMGEXP**) are encoded as text, and irregularly.  We translate these variables to numbers, and then we compute the damages in dollars and store it in the **PROPDMG** and **CROPDMG** variables, respectively.
```{r}
dictPROPDMGEXP <- c(0,0,0,0,0,1,2,3,4,5,6,7,8,9,2,2,3,6,6)
names(dictPROPDMGEXP) <- levels(x$PROPDMGEXP)
dictCROPDMGEXP <- c(0,0,0,2,9,3,3,6,6)
names(dictCROPDMGEXP) <- levels(x$CROPDMGEXP)
x$PROPDMGEXP <- sapply(x$PROPDMGEXP, function(z) return(dictPROPDMGEXP[z]))
x$CROPDMGEXP <- sapply(x$CROPDMGEXP, function(z) return(dictCROPDMGEXP[z]))
x$PROPDMG <- x$PROPDMG * 10^(x$PROPDMGEXP)
x$CROPDMG <- x$CROPDMG * 10^(x$CROPDMGEXP)
```
Many of the event types are similar, and in particular, may differ only by minor spelling or capitalization issues in the variable name.  Thus we wish to condense our variables into broader categories.

We write two functions, the first of which is

```{r}
condense <- function(yy,expr,newexpr) {
  y <- yy
  y$EVTYPE <- as.factor(gsub(expr,newexpr,y$EVTYPE))
  y
}
```
which relabels all event types meeting a certain pattern (regular expression **expr**) with a specified new event category (**newexpr**).  We also write a utility function **peek** to assist with the (somewhat painstaking) process of condensing the category types.
```{r}
peek <- function(data,expr) {
  if (expr == "") return(levels(data$EVTYPE))
  else return(levels(data$EVTYPE)[grepl(expr,levels(data$EVTYPE))])
}
```
(To clarify: **peek** was used in the RStudio Console to help determine the expressions used in the next code block.  So although it does not appear explicitly, it was quite helpful in writing the code that is used.)

At this point we are ready to condense the event types to a few major categories.  We proceed sequentially.  Some of the expressions below are rather complex; they were written by repeatedly using **peek** to determine categories to gather together.
```{r}
y <- x
y <- y[!grepl(".*Summary.*|.*SUMMARY.*|.*RECORD.*|.*Record.*|.*VOLCANIC.*|.*Volcanic.*",y$EVTYPE),]
y$EVTYPE <- droplevels(y$EVTYPE)
y <- condense(y,".*FIRE.*|.*SMOKE.*","Fire")
y <- condense(y,".*TORN.*|.*Tornado.*|.*FUNNEL.*|.*Funnel.*","Tornado")
y <- condense(y,".*LANDSLI.*|.*AVALAN.*|.*Landsli.*|.*ROCK S.*|.*Mudslide.*|.*MUDSLIDE.*|.*MUD SLIDE.*","Landslide,Avalanche")
y<- condense(y,".*MARINE.*|.*Marine.*|.*HURRICANE.*|.*Hurricane.*|.*COASTAL.*|.*coastal.*|.*SURF.*|.*Surf.*|.*surf.*|.*Beach.*|.*Coastal.*|.*SEAS.*|.*TIDAL.*|.*Tidal.*|.*CURRENT.*|.*TYPHOON.*|.*ROGUE WAVE.*|.*TSUNAM.*|.*TROPICAL.*|.*STORM SURGE.*|.*FOG.*|.*WATERSPOUT.*|.*HIGH WAVES.*|.*SWELLS.*|ASTRONOMICAL HIGH TIDE|ASTRONOMICAL LOW TIDE|.*HEAVY SWELLS.*|.*HIGH TIDES.*|.*FLOYD.*|.*BLOW-OUT.*","Marine,Coastal")
y <- condense(y,".*THUND.*|.*Thund.*|.*LIGHTN.*|.*THUNER.*|.*THUDER.*|LIGNTNING|LIGHTING|.*TSTM.*","Lightning,Thunderstorms")
y <- condense(y,".*HEAT.*|.*Heat.*|.*DROUGHT.*|.*Drought.*|.*WARM.*|.*WARM WEATHER.*|.*RECORD HIGH.*|.*HOT.*|.*Hot.*","Heat,Drought")
y <- condense(y,".*ICE.*|.*Ice.*|.*ICY.*|.*Icy.*|.*WINT.*|.*SNOW.*|.*Snow.*|.*snow.*|.*Wint.*|.*COLD.*|.*Cold.*|.*FREEZ.*|.*Freez.*|.*BLIZZ.*|.*CHILL.*|.*MIXED PRECIP.*|.*HYPOTHERMIA.*|.*Hypothermia.*|.*HYPERTHERM.*|.*SLEET.*|.*Frost.*|.*FROST.*|HEAVY MIX|Ice jam flood.*|.*LOW TEMPERATURE.*|.*Mixed Precipitation.*","Winter Conditions")
y <- condense(y,".*RAIN.*|.*Rain.*|.*rain.*|.*FLOOD.*|.*URBAN.*|.*Flood.*|.*HEAVY PRECIPITATION.*|.*HEAVY PRECIPATATION.*|HEAVY SHOWER|HIGH WATER|DOWNBURST|RAPIDLY RISING WATER|.*DAM.*|.*SMALL STREAM.*|.*Sml Stream.*","Rain,Flooding")
y <- condense(y,".*HAIL.*|.*Hail.*|.*hail.*","Hail")
y <- condense(y,".*WIND.*|.*Wind.*|.*wind.*|.*GUST.*|.*DEVIL.*|.*Devil.*|.*DUST.*|Microburst|.*MICROBURST.*|SEVERE TURBULENCE","Winds")

y <- condense(y,"APACHE COUNTY|COOL AND WET|DENSE SMOKE|DROWNING|DRY MICROBURST|EXCESSIVE WETNESS|Glaze|GLAZE|HIGH|Landslump|LANDSPOUT|OTHER|SEICHE|WET Winds|\\?","Other")

y <- y[!grepl(".*ABNORMALLY.*|.*BELOW.*|.*RECORD.*|.*Record.*|.*BEACH.*|.*NORTHERN.*|.*No S.*|.*DRIEST.*|.*EXCESSIVE.*|Mild.*|MILD.*|.*COOL SP.*|.*LANDSLUM.*|.*MONTHLY.*|.*NORMAL.*|.*Metro.*|.*EXTREMELY WET.*|.*Saharan.*|.*VOG.*|.*RED FLAG.*|.*NONE.*|.*ROTATING.*|.*SOUTHEAST.*|.*CLOUD.*|.*WET.*|.*Wet.*|.*wet.*|.*WND.*|.*record.*",y$EVTYPE),]
y$EVTYPE <- droplevels(y$EVTYPE)

y <- condense(y,".*DRY.*","Heat,Drought")
y <- condense(y,".*FLOOOD.*|.*Precip.*|.*Urban.*|.*dingS.*","Rain,Flooding")
y<- condense(y,".*WAYTER.*|.*SPOUT.*","Marine,Coastal")
```
Now that we have condensed our data set to a few broad categories, we provide a summary table for the data (displayed in the **Results** section).
```{r}
summarytable <- as.data.frame(sapply(y[,sapply(y,is.numeric)],
                       function(z) by(z,y$EVTYPE,sum)))
names(summarytable) <- c("Fatalities","Injuries","Property Damage (US $)","Crop Damage (US $)")

```

## Results ##
We first summarize the results in two tables.
 
The effect of severe weather as measured by human injury and fatality:
```{r}
print(summarytable[,c("Injuries","Fatalities")])
```
The effect of severe weather as measured by property and crop damage:
```{r}
print(summarytable[,c("Property Damage (US $)","Crop Damage (US $)")])
```
The human cost (injury/fatality) of severe weather in the US is shown in the following plot.
```{r}
injury<-summarytable$Injuries
names(injury)<-row.names(summarytable)
fatality <- summarytable$Fatalities
names(fatality) <- row.names(summarytable)
nall<-row.names(summarytable)
nsort <- names(sort(injury[nall]+fatality[nall],decreasing=FALSE))
injury <- injury[nsort]
fatality <- fatality[nsort]
results <- rbind(injury,fatality)/1000
par(mai=c(1,3,1,1),mfrow=c(1,1))
barplot(results,horiz=TRUE,las=1,col=c("darkblue","red"),
        beside=FALSE,legend.text=c("Injuries","Fatalities"),
        args.legend = list(x=95,y=5),
        main="Injuries/Fatalities for Severe Weather",
        xlab = "Injuries/fatalities in thousands")
```

The economic cost (property/crop damage) of severe weather is shown in the following plot.
```{r}
crop<-summarytable[,"Crop Damage (US $)"]
names(crop) <- row.names(summarytable)
prop<-summarytable[,"Property Damage (US $)"]
names(prop) <- row.names(summarytable)
crop <- crop[nsort]
prop <- prop[nsort]
results <- rbind(prop/100000000,crop/1000)
par(mai=c(1,3,1,1),mfrow=c(1,1))
barplot(results,horiz=TRUE,las=1,col=c("purple4","green4"),
        beside=TRUE,legend.text=c("Property Damage","Crop Damage"),
        args.legend = list(x=1600,y=10),
        main="Property/Crop Damage for Severe Weather",
        xlab = "Property Damage, Hundreds of Millions of U.S. $\nCrop Damage, U.S. $",ylab="")
```

## Tornado Damage ##
Because tornadoes seem to pose a much greater threat to human health, and because the majority of tornado damage is limited to certain geographic regions, we provide a separate analysis of tornado event data.

We begin by extracting only the tornado event data and displaying the table of the 20 states with the highest (total) fatalities due to tornado events.  (The remaining variables are roughly correlated with fatalities, and so this gives a fair overall sense of the states "hit hardest" by tornadoes.)
```{r}
t <- y[y$EVTYPE=="Tornado",c(1,2,3,4,6,8,9,10,11)]
ttable <- as.data.frame(sapply(t[,sapply(t,is.numeric)],
                       function(z) by(z,t$STATE,sum)))
names(ttable) <- c("Fatalities","Injuries","Property Damage (US $)","Crop Damage (US $)")
ttable[order(ttable$Fatalities,decreasing=TRUE),][1:20,]
```

We also note that injuries and fatalities caused by tornadoes has changed over time, and differs from month to month.  In the following panel plot,
```{r,fig.width=8,fig.height=6}
timetable <- as.data.frame(sapply(t[,sapply(t,is.numeric)],
                       function(z) by(z,t$YEAR,sum)))
dates <- as.numeric(as.character(row.names(timetable)))
monthtable <- as.data.frame(sapply(t[,sapply(t,is.numeric)],
                       function(z) by(z,t$MONTH,sum)))
monthtable <- monthtable[,c("INJURIES","FATALITIES")]
par(mfrow=c(1,2))
plot(dates,timetable$FATALITIES,type="l",col="red",
     xlab="Year",ylab="Injuries/Fatalities",
     ylim=c(0,max(timetable$FATALITIES,timetable$INJURIES)))
par(new=TRUE)
plot(dates,timetable$INJURIES,type="l",col="blue",
     xlab="",ylab="",
     ylim=c(0,max(timetable$FATALITIES,timetable$INJURIES)))
barplot(t(monthtable)[,c("Jan","Feb","Mar","Apr","May",
                         "Jun","Jul","Aug","Sep","Oct","Nov","Dec")],
        col=c("darkblue","red"),beside=FALSE,
        legend.text=c("Injuries","Fatalities"),
        args.legend = list(x=15,y=30000),main="",
        xlab="Month",ylab="Injuries/Fatalities")
```

we see that injuries/fatalities seem to decrease in variance over time, and in particular, the number of spikes decreasing over time.  This seems to be consistent with our intuition about the effects of technological advances (better communication and Doppler radar, primarily) on our ability to provide early warnings for tornadoes.  We can also see in the graph that the relative rarity of tornadoes means that a single severe event may be enough to cause a spikein injuries/fatalities, even with modern technology.  Consider, for instance, the spike in the injury/fatality graph in 1999; this corresponds to [the events occurring May 2-8, 1999](http://en.wikipedia.org/wiki/1999_Oklahoma_tornado_outbreak), during which (May 3) an F5 class tornado struck the Oklahoma City area (and incidentally, this was the first use of the new "tornado emergency" system).  For comparison, the following table shows the injuries and fatalities in the entire U.S. in the year 1999, compared to the injuries and fatalities reported in the May 3, 1999 events in Oklahoma.

```{r}
oneyear<-y[which(y$EVTYPE == "Tornado" & grepl(".*1999.*",y$BGN_DATE)),]
ok<-oneyear[which(grepl("5/[2,3,4,5,6,7,8]/1999.*",oneyear$BGN_DATE) & oneyear$STATE == "OK"),]
oktable <- sapply(ok[,sapply(ok,is.numeric)],sum)
oneyeartable <- sapply(oneyear[,sapply(oneyear,is.numeric)],sum)
ptable<-paste(formatC(oktable/oneyeartable * 100,width=3),"%",sep="")
maytable <- rbind(oktable[c("INJURIES","FATALITIES")],oneyeartable[c("INJURIES","FATALITIES")],ptable[c(1,2)])
row.names(maytable) <- c("Oklahoma City OK, May 2-8, 1999","U.S., entire year 1999","Percentage caused by May OK event")
maytable
```

As we see, the May 3 event makes up a large percentage of the injuries and fatalities for the entire year, accounting for the spike seen in the injuries/fatalities over time plot.  This demonstrates that these plots can be sensitive to single, extreme events; unsurprising, but worth noting.

## Conclusions ##
Severe weather causes injury, death, and economic damage across the U.S. every year.  We have provided a categorization of the events listed in the NOAA Severe Weather data, and we have summarized damages caused by those broad categories.  Finding that tornadoes account for a comparatively large amount of damages, we investigate tornado damages further.  We find that over time, injuries and fatalities caused by tornadoes in the U.S. have decreased, but that these findings can be skewed easily by extreme events.  We also found that over time, the property and crop damages caused by tornadoes have increased.  This result is questionable, as the damages were not adjusted for inflation.
